Parallel Monte Carlo Synthetic Acceleration Methods for Discrete Transport Problems

### 2013-08-28 doc/dissertation/presentation/defense.tex
apreliminary updates to defense talk

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}
  \begin{itemize}
  \item Development of a linear scheme for discrete systems leveraging
    Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Application to neutron transport
    \item Research is required to explore general MCSA development
    \item Performance is of concern
    \end{itemize}
    \bigskip
  \item Development of a nonlinear scheme for discrete systems
    leveraging Monte Carlo Synthetic Acceleration
    \begin{itemize}
    \item Monte Carlo is a more natural fit
    \item Application to fluid flow
    \item Convergence of the linear model is of concern
    \end{itemize}
    \bigskip
  \item Parallelization of Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Parallel strategies taken from modern reactor physics
      methods
    \item Research is required to explore varying parallel strategies
    \item Scalability is of concern
    \end{itemize}
  \end{itemize}
\end{frame}

@@

@@

  \begin{itemize}
  \item Swap in MCSA as the solver at each eigenvalue operation
    \medskip
  \item Use Thyra/Stratimikos interfaces for plug and play with $SP_N$
    \medskip
  \item Current Exnihilo implementation gives the full operator
  \end{itemize}

    \begin{column}{0.5\textwidth}
      {\small
        \begin{itemize}
        \item CASL Problem 3: $17 \times 17$ quarter symmetry HZP LWR
          fuel assembly
        \item $SP_N$ criticality calculation
        \item MCLS leveraged by the Exnihilo code base (ORNL)
        \item Comparison to Trilinos Aztec Krylov solvers with ILUT
        \end{itemize}
      }
      {\tiny
        \begin{table}[htpb!]
          \begin{center}
            \begin{tabular}{ll}\hline\hline
              \multicolumn{1}{l}{\textbf{Parameter}} & 
              \multicolumn{1}{l}{\textbf{Value}} \\
              Power Level & 0 MW \\
              Inlet Temperature & 326.85C \\
              Fuel Temperature & 600C \\
              Boron Concentration & 1300 ppm \\
              Moderator Density & 0.743 g/cc \\
              Helium Density & \sn{1.79}{-4} g/cc \\
              Zirconium Density & 6.56 g/cc \\
              Stainless Steel Density & 8.0 g/cc \\
              Inconel Density & 8.19 g/cc \\
              UO2 Density & 10.257 g/cc \\
              Fuel Pin Radius (w/o clad) & 0.4096 cm \\
              %%
              \hline\hline
            \end{tabular}
          \end{center}
        \end{table}
      }
    \end{column}

@@

    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Light water moderator creates a lot of scattering and
        $\mathbf{\rho}(\mathbf{H}) \approx 1$
        \medskip
      \item Intractable number of histories required for MCSA
        convergence 
      \end{itemize}
    \end{column}

@@

@@

      \begin{itemize}
        \item MCLS generates the same k-eigenvalue and neutron flux in
          all groups and spatial locations
          \bigskip
        \item MCSA converged in fewer iterations than GMRES, more
          iterations than BiCGStab
          \bigskip
        \item Explicit preconditioning strategy destroys sparsity and
          elevates CPU times (Ifpack ILUT)
          \bigskip
        \item Spectral radius and memory limitations combine to
          prevent solutions at finer discretizations
      \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Synthetic Acceleration for Nonlinear Problems}

  \begin{itemize}
  \item Many physics problems of interest are nonlinear
    \bigskip
  \item Signficant research on Newton methods since the 1980's
    \bigskip
  \item Newton methods often leverage Krylov solvers
    \begin{itemize}
    \item Robust implementations
    \item No operator required
    \end{itemize}
    \bigskip
  \item Monte Carlo methods need the full operator
    \bigskip
  \item Automatic construction of the linear model is available
    \begin{itemize}
    \item Operator overloading for nonlinear residual differentiation
    \item Ideal for Monte Carlo
    \item Potential scaling improvements
    \item Potential resiliency benefits
    \end{itemize}
  \end{itemize}

@@

  \begin{itemize}
  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
    \begin{itemize}
    \item Thermal convection cavity problem (De Vahl Davis, 1983)
    \item Lid driven cavity problem (Ghia et al., 1982)
    \item Backward-Facing step problem (Gartling, 1990)
    \end{itemize}
    \medskip
  \item Tuning benchmark parameters varies the strength of
    nonlinearities
    \medskip
  \item Newton-Krylov method leveraging Aztec GMRES used for
    comparisons
    \medskip
  \item All problems preconditioned with algebraic multigrid (ML) and
    leveraged some kind of globalization (e.g. backtracking)
  \end{itemize}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{drekar1.png}
        \end{center}
      \end{figure}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{drekar2.png}
        \end{center}
      \end{figure}
    \end{column}
  \end{columns}

  \begin{itemize}
  \item Drekar is a production physics suite that leverages Newton
    methods with FAD
  \bigskip
  \item MCLS incorporated into the Drekar nonlinear solver scheme
    using Thyra/Stratimikos to implement FANM
  \end{itemize}

@@

  \begin{figure}[htpb!]
    \begin{center}
      \includegraphics[width=4in]{titan_pure_strong.pdf}
    \end{center}
    \caption{\textbf{Pure domain decomposition.} \textit{Super-linear
        speed-up from memory thrashing in base case. MCLS is an order
        of magnitude slower arithmetically.}}
  \end{figure}

@@

  \begin{columns}
    \begin{column}{0.55\textwidth}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{titan_weak_absolute.pdf}
        \end{center}
      \end{figure}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.5in]{titan_weak_implementation.pdf}
        \end{center}
      \end{figure}

@@

    \begin{column}{0.45\textwidth}

      \begin{itemize}
      \item CG demonstrates poor scaling due to the cheaper iteration
        sequence
        \bigskip
      \item Implementation efficiency is effectively the parallel
        efficiency of a single iteration
        \bigskip
      \item Improved implementation efficiency means improved
        iterative performance will potentially give a better time to
        solution
      \end{itemize}

@@

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Strong Scaling Results with Overlap}

@@

  \small{
    \begin{itemize}
    \item MCSA has been incorporated into the Exnihilo neutronics
      production code base developed at Oak Ridge National Laboratory
      \medskip
    \item MCSA can solve the asymmetric system generated by the $SP_N$
      equations
      \medskip
    \item Light water reactor problems are difficult to solve with MCSA as
      they have large spectral radii due to the neutron scattering in the
      moderator
      \medskip
    \item Advanced algebraic preconditioning strategies were applied to
      the $SP_N$ equations to obtain convergence with ILUT chosen for
      subsequent investigations
      \medskip
    \item MCSA was observed to converge in fewer iterations per eigenvalue
      iteration than GMRES for the fuel assembly criticality problem and
      more than Bi-CGStab using the same preconditioning
    \end{itemize}
  }

@@

  \small{
    \begin{itemize}
    \item Forward-Automated Newton-MCSA (FANM) has been developed
      \medskip
    \item The FANM method has been incorporated into the Drekar
      multiphysics production code base developed at Sandia National
      Laboratories
      \medskip
    \item The FANM method has better iterative performance than the
      Newton-Krylov method for convection dominated problems, converging
      in fewer linear solver iterations with the same preconditioning for
      high and low Rayleigh numbers
      \medskip
    \item The spectral radius convergence restriction on MCSA was observed
      to be a significant hindrance by preventing solutions to forced flow
      problems at high Reynolds numbers
      \medskip
    \item More Monte Carlo histories at every FANM iteration can reduce
      the number of linear and nonlinear iterations required to converge
      the problem
    \end{itemize}
  }

@@

  \begin{itemize}
  \item Shortcomings observed on real problems
    \begin{itemize}
    \item Significant optimization required to determine production
      feasibility and true scalability
    \item Explicit algebraic preconditioning methods not sufficient
    \item Spectral radius limitation is severe
    \end{itemize}
    \medskip
  \item Performance improvements
    \begin{itemize}
    \item Random walk optimizations
    \item Alias sampling implementations
    \item Multiple set reduction analysis
    \item FANM forcing term and MCSA history relationships
    \end{itemize}
    \medskip
  \item Preconditioning improvements
    \begin{itemize}
    \item Alternative left/right sampling schemes
    \item Variance reduction based strategy
    \item Reduced order physics/PDE models for acceleration
    \end{itemize}
    \medskip
  \item Breaking away from $\rho(\mathbf{H}) < 1$
    \begin{itemize}
    \item Monte Carlo methods of the second degree
    \item Stochastic projection methods
    \end{itemize}
  \end{itemize}

@@

  \small{
    \begin{enumerate}
    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A
      Multiple-Set Overlapping-Domain Decomposed Monte Carlo Synthetic
      Acceleration Method for Linear Systems}, \textit{Joint
      International Conference on Supercomputing in Nuclear
      Applications and Monte Carlo 2013 (SNA+MC 2013), Paris, France,
      October 27-31, 2013. Accepted for oral presentation.}
      \medskip
    \item T.M. Evans, S.W. Mosher, S.R. Slattery, S.P. Hamilton,
      \textbf{A Monte Carlo Synthetic-Acceleration Method for Solving
        the Thermal Radiation Diffusion Equation}, \textit{Journal of
        Computational Physics, Submitted.}
      \medskip
    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A Spectral
      Analysis of the Domain Decomposed Monte Carlo Method for Linear
      Systems}, \textit{International Conference on Mathematics and
      Computational Methods Applied to Nuclear Science \& Engineering
      (M\&C 2013), American Nuclear Society, Sun Valley, ID, May 5-9,
      2013.}
    \end{enumerate}
  }
\end{frame}

@@

  \begin{itemize}
  \item Paul
  \item Tom
  \item Roger
  \item CASL, ORNL, Sandia, OLCF
\end{itemize}

>>>>

%%---------------------------------------------------------------------------%%
\begin{frame}{Physics-Based Motivation}

  \begin{itemize}
  \item Physics-driven development
    \begin{itemize}
    \item Identify the physics of interest and research MCSA
      applicability
    \item Offer solutions or a potential path forward to the observed
      issues
    \end{itemize}
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}
  \begin{itemize}
  \item Development of a linear scheme for discrete systems leveraging
    Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Application to neutron transport
    \item Research is required to study MCSA preconditioning
    \item Iterative performance is of concern
    \end{itemize}
    \bigskip
  \item Development of a nonlinear scheme for discrete systems
    leveraging Monte Carlo Synthetic Acceleration
    \begin{itemize}
    \item Monte Carlo is a more natural fit
    \item Application to fluid flow
    \item Convergence of the linear model is of concern
    \end{itemize}
    \bigskip
  \item Parallelization of Monte Carlo Synthetic Acceleration
    \medskip
    \begin{itemize}
    \item Parallel strategies taken from modern reactor physics
      methods
    \item Research is required to explore varying parallel strategies
    \item Parallel scalability is of concern
    \end{itemize}
  \end{itemize}
\end{frame}

@@

    \begin{column}{0.5\textwidth}
      {\small
        \begin{itemize}
        \item CASL Problem 3: $17 \times 17$ quarter symmetry HZP LWR
          fuel assembly
        \item Multigroup $SP_N$ discretization
        \item MCLS leveraged by the Exnihilo code base (ORNL)
        \item Emphasize algorithm development for iterative
          performance
        \end{itemize}
      }
      {\tiny
        \begin{table}[htpb!]
          \begin{center}
            \begin{tabular}{ll}\hline\hline
              \multicolumn{1}{l}{\textbf{Parameter}} & 
              \multicolumn{1}{l}{\textbf{Value}} \\
              Power Level & 0 MW \\
              Inlet Temperature & 326.85C \\
              Fuel Temperature & 600C \\
              Boron Concentration & 1300 ppm \\
              Moderator Density & 0.743 g/cc \\
              Helium Density & \sn{1.79}{-4} g/cc \\
              Zirconium Density & 6.56 g/cc \\
              Stainless Steel Density & 8.0 g/cc \\
              Inconel Density & 8.19 g/cc \\
              UO2 Density & 10.257 g/cc \\
              Fuel Pin Radius (w/o clad) & 0.4096 cm \\
              %%
              \hline\hline
            \end{tabular}
          \end{center}
        \end{table}
      }
    \end{column}

@@

  \begin{itemize}
  \item Inject MCSA as the solver at each eigenvalue operation
    \medskip
  \item Transport operator is static - one time cost for MCSA setup
    \medskip
  \item Current Exnihilo implementation gives the full operator
  \end{itemize}

@@

    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Rapidly divergent results
        \medskip
      \item Light water moderator creates a lot of scattering and
        $\mathbf{\rho}(\mathbf{H}) \approx 1$
        \medskip
      \item Convergence not achieved with 50 histories per DOF and 90
        minutes compute time
      \end{itemize}
    \end{column}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{ILUT Preconditioning}

      \begin{figure}[t!]
        \begin{center}
          \includegraphics[width=2.0in]{ilut_iterations.pdf}
        \end{center}
      \end{figure}

      \begin{figure}[t!]
        \begin{center}
          \includegraphics[width=2.0in]{ilut_size.pdf}
        \end{center}
      \end{figure}

      \begin{itemize}
      \item {\small Factor transport operator into upper and lower
        triangular parts}
        \begin{itemize}
        \item {\small $\mathbf{R} = \mathbf{L} \mathbf{U} -
          \mathbf{M}$}
        \end{itemize}
      \item {\small Control factorization content with level-of-fill
        and drop tolerance}
      \item {\small Use the reduced domain approximation to recover
        sparsity}
      \end{itemize}

      \begin{figure}[t!]
        \begin{center}
          \includegraphics[width=2.0in]{rda_ilut_iterations.pdf}
        \end{center}
      \end{figure}

@@

      \begin{itemize}
        {\small
        \item Comparison to Trilinos Aztec Krylov solvers with ILUT
          \medskip
        \item MCLS generates the same k-eigenvalue and neutron flux in
          all groups and spatial locations
          \medskip
        \item MCSA converged in fewer iterations than GMRES, more
          iterations than BiCGStab
          \medskip
        \item Explicit preconditioning strategy destroys sparsity and
          elevates CPU times (Ifpack ILUT)
          \medskip
        \item Spectral radius and memory limitations combine to
          prevent solutions at finer discretizations
        }
      \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Navier-Stokes Benchmark Problems}

  \begin{itemize}
  \item Sequence of Navier-Stokes benchmarks for the nonlinear methods
    \begin{itemize}
    \item Thermal convection cavity problem (De Vahl Davis, 1983)
    \item Lid driven cavity problem (Ghia et al., 1982)
    \item Backward-Facing step problem (Gartling, 1990)
    \end{itemize}
    \medskip
  \item Tuning benchmark parameters varies the strength of
    nonlinearities
    \medskip
  \end{itemize}

%%----------s-----------------------------------------------------------------%%
\begin{frame}{MCSA for Nonlinear Problems}

  \begin{itemize}
  \item Need a nonlinear solution scheme for the Navier-Stokes equations
    \bigskip
  \item Signficant research on Newton methods since the 1980's
    \bigskip
  \item Newton methods often leverage Krylov solvers
    \begin{itemize}
    \item Robust implementations
    \item No operator required
    \end{itemize}
    \bigskip
  \item Monte Carlo methods need the full operator
    \bigskip
  \item Automatic construction of the linear model is available
    \begin{itemize}
    \item Operator overloading for nonlinear residual differentiation
    \item Ideal for Monte Carlo
    \item Potential parallel scaling and iterative performance
      improvements
    \item Potential resiliency benefits
    \end{itemize}
  \end{itemize}

@@

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.0in]{drekar1.png}
        \end{center}
      \end{figure}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.0in]{drekar2.png}
        \end{center}
      \end{figure}
    \end{column}
  \end{columns}

  \medskip

  \begin{itemize}
    {\small
    \item Drekar is a production fluid code base using Newton methods
      with FAD
      \medskip
    \item MCLS incorporated into Drekar nonlinear scheme to implement
      FANM
      \medskip
    \item Newton-Krylov method with Aztec GMRES used for benchmark
      comparisons
    \item All problems preconditioned with explicit scheme using
      algebraic multigrid (ML) and leveraged some kind of
      globalization (e.g. backtracking) 
    }
  \end{itemize}

@@

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.4in]{titan_pure_strong_time.pdf}
        \end{center}
        \caption{CPU Time.}
      \end{figure}

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.4in]{titan_pure_strong.pdf}
        \end{center}
        \caption{Absolute Efficiency.}
      \end{figure}

  \begin{itemize}
  \item MCLS is an order of magnitude slower arithmetically.
  \item Super-linear speed-up from memory thrashing in base case.
  \item CG demonstrates poor scaling due to the cheaper iteration
    sequence
  \end{itemize}

@@

      \begin{figure}[htpb!]
        \begin{center}
          \includegraphics[width=2.4in]{titan_pure_weak_time.pdf}
        \end{center}
        \caption{CPU Time.}
      \end{figure}

@@

    \begin{figure}[htpb!]
      \begin{center}
        \includegraphics[width=2.4in]{titan_weak_absolute.pdf}
      \end{center}
      \caption{Absolute Efficiency.}
    \end{figure}

@@

  \begin{itemize}
  \item Improved implementation efficiency means improved
    iterative performance will potentially give a better time to
    solution
    \bigskip
  \item Implementation efficiency is effectively the parallel
    efficiency of a single iteration
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Scaling Results with Overlap}

@@

  \small{
    \begin{itemize}
    \item MCSA has been incorporated into the Exnihilo neutronics
      production code base developed at Oak Ridge National Laboratory
      \medskip
    \item MCSA can solve the asymmetric system generated by the $SP_N$
      equations
      \medskip
    \item Light water reactor problems are difficult to solve with MCSA as
      they have large spectral radii due to the neutron scattering in the
      moderator
      \medskip
    \item Advanced algebraic preconditioning strategies were applied to
      the $SP_N$ equations to obtain convergence with ILUT chosen for
      subsequent investigations
      \medskip
    \item MCSA with the reduced domain approximation was observed to
      converge in fewer iterations per eigenvalue iteration than GMRES
      for the fuel assembly criticality problem and more than
      Bi-CGStab using the same preconditioning
    \end{itemize}
  }

@@

  \small{
    \begin{itemize}
    \item Forward-Automated Newton-MCSA (FANM) has been developed
      \medskip
    \item The FANM method has been incorporated into the Drekar
      multiphysics production code base developed at Sandia National
      Laboratories
      \medskip
    \item The FANM method has been verified to produce the same solutions
      as a production Newton-Krylov method for three difficult benchmark
      problems for the Navier-Stokes equations in different flow regimes
      and geometries
      \medskip
    \item The FANM method has better iterative performance than the
      Newton-Krylov method for convection dominated problems, converging
      in fewer linear solver iterations with the same preconditioning for
      high and low Rayleigh numbers
      \medskip
    \item The spectral radius convergence restriction on MCSA was observed
      to be a significant hindrance by preventing solutions to forced flow
      problems at high Reynolds numbers
    \end{itemize}
  }

@@

  \begin{itemize}
  \item Shortcomings observed on real problems
    \begin{itemize}
    \item Significant optimization required to determine production
      feasibility and true scalability
    \item Explicit algebraic preconditioning methods not sufficient
    \item Spectral radius limitation is severe
    \end{itemize}
    \medskip
  \item Performance improvements
    \begin{itemize}
    \item Random walk optimizations
    \item Multiple set reduction analysis
    \item FANM forcing term and MCSA history relationships
    \end{itemize}
    \medskip
  \item Preconditioning improvements
    \begin{itemize}
    \item Variance reduction based strategy
    \item Reduced order physics/PDE models for acceleration
    \end{itemize}
    \medskip
  \item Breaking away from $\rho(\mathbf{H}) < 1$
    \begin{itemize}
    \item Monte Carlo methods of the second degree
    \end{itemize}
  \end{itemize}

@@

  \small{
    \begin{enumerate}
    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A
      Multiple-Set Overlapping-Domain Decomposed Monte Carlo Synthetic
      Acceleration Method for Linear Systems}, \textit{Joint
      International Conference on Supercomputing in Nuclear
      Applications and Monte Carlo 2013 (SNA+MC 2013), Paris, France,
      October 27-31, 2013. Accepted for oral presentation.}
      \medskip
    \item S.R. Slattery, T.M. Evans, P.P.H. Wilson, \textbf{A Spectral
      Analysis of the Domain Decomposed Monte Carlo Method for Linear
      Systems}, \textit{International Conference on Mathematics and
      Computational Methods Applied to Nuclear Science \& Engineering
      (M\&C 2013), American Nuclear Society, Sun Valley, ID, May 5-9,
      2013.}
      \medskip
    \item T.M. Evans, S.W. Mosher, S.R. Slattery, S.P. Hamilton,
      \textbf{A Monte Carlo Synthetic-Acceleration Method for Solving
        the Thermal Radiation Diffusion Equation}, \textit{Journal of
        Computational Physics, Submitted.}
    \end{enumerate}
  }
\end{frame}

@@

  \begin{itemize}
  \item Paul Wilson
  \item Tom Evans
  \item Roger Pawlowski
  \item CASL, ORNL, Sandia, OLCF
  \end{itemize}
